Spatial Reasoning and Data Displays

### 2015-06-08 Paper/InfoVis2015/ReviewerResponses.Rmd

---
title: "ReviewerResponses"
author: "Susan VanderPlas"
date: "June 8, 2015"
output: html_document
---

# Summary Rating

   4  (Accept) The paper should be accepted with some minor revisions. Once these have been completed it will meet the quality standard.

## The Summary Review

### Overview

   The paper attempts to solve a quite interesting problem, which is
   understanding the relationship between demographic characteristics /
   visual skills and perception of graphical lineups. Reviewers agree that
   this topic is quite interesting, and the authors did conduct a quite
   extensive study. R4, particularly, is strongly supportive for this paper,
   saying "The work represented here may be the most important recent
   advance in methodology for evaluating the effectiveness of visual
   displays." However, other reviewers listed several concerns in
   experimental design, data analysis, and conclusions.

### Strengths

   - The topic of this study is very interesting. (R3, R4)
   - The paper is well presented. (R1, R2, R3, R4)
   - The study is extensively evaluated. (R1)
   - Supplementary materials provide additional details on methodologies
   used. (R2)
   - A groundbreaking methodological contribution is made. (R4)
   - Deal with important issues in spatial reasoning on graphics. (R1)

### Weaknesses

   - A clearer motivation of certain experimental design choices is
   necessary. (R1)
   - Experiment procedures should have more proper control and be more
   detailed. (R1, R3)
   - The statistical analysis is of questionable validity. (R2, R3)
   - Results are not conclusive and open to interpretation. (R1, R2)
   - A lack of discussion on visualization design in general. (R3)
   - A separate related work section is necessary. (R3)

   However, these concerns, especially regarding statistical analysis, are
   rather critical. The reviewers recommended to accept this paper in the
   condition that these weaknesses are properly handled within the revision
   cycle. Thus, as the primary reviewer of this paper, I will review the
   revision of the paper and ask the following questions, and if the answers
   to all the questions are "yes", I will recommend to accept this paper.

   Q1. Are all the experimental design choices properly motivated? (refer to
   the first concern of R1).
   Q2. Are the experimental procedure sufficiently detailed? (refer to the
   second concern of R1 and )
   Q3. The results of statistical analysis do not seem to be conclusive, are
   the interpretations of these results properly revised? (refer to the
   third concern of R1)
   Q4. Does the revision contain additional discussion of why PCA is a
   proper approach to interpret the data for the VIS community members who
   do not have sufficient statistics background? (refer to the concern of
   R2)
   Q5. Does the revision has the clear description of "most different"?
   (refer to the first concern of R3)
   Q6. Does the revision has the proper discussion of visualization design
   (refer to the second concern of R3)
   Q7. Does the revision resolve a potential concern of confounding factors
   (refer to the conern of R3)
   Q8. Does the revision have an explicit related works section? (refer to
   the last concern of R3)
   Q9. Does the revision handle all the minor issues raised by reviewers?

Second round comments (public)

Second round supplementary materials check

Second round supplementary materials comments

# Review 2
Title: Spatial Reasoning and Data Displays

_Reviewer_: primary

_Paper type_   Evaluation

_Expertise_ 2  (Knowledgeable)

_Overall Rating_ 3.5 - Between Possible Accept and Accept

_Supplemental Materials_ Acceptable

## Justification

   Overall, the study appears to be meticulously conducted, and the research
   question of "what kinds of factors really influence on visual tasks" is a
   very needed to be scientifically answered. However, I have a concern
   whether PCA is a proper inferencing tool (I think it as a more
   dimensionality reduction tool) and the resulting take-home messages are
   rather unclear.

## The Review

   Overall, various statistical methods (e.g., scaling scores) appeared to
   be meticulously used, and the authors even provided supplementary
   materials to detail such processes, which shows a great practice. Using a
   previously tested approach, line-ups, to investigate which factors will
   impact on visualization task performances is also an interesting idea. I
   believe that our community needs more work like this.

   However, I have a concern. The authors largely rely on principal
   component analysis (PCA) to infer what kinds of cognitive abilities are
   relevant with the line-up tasks. However, I wonder if that is a
   statistically sound approach. I am not a statistician, so I cannot make a
   definitive claim that this is a right or wrong approach (I hope that
   other reviewers have a stronger background on this approach).

   However, I found the following page
   (http://www.alglib.net/dataanalysis/principalcomponentsanalysis.php)
   describing the underlying assumptions of PCA:

   "It may be noted that the PCA is based on the following assumptions:

   The assumption that the dimensionality of data can be efficiently reduced
   by linear transformation
   The assumption that most information is contained in those directions
   where input data variance is maximum.
   As it is evident, these conditions are by no means always met. For
   example, if points of an input set are positioned on the surface of a
   hypersphere, no linear transformation can reduce dimension (nonlinear
   transformation, however, can easily cope with this task). This
   disadvantage is equally attributable to all linear algorithms, and it can
   be eliminated due to the use of complementary dummy variables that are
   nonlinear functions of the input data set elements (the so-called "kernel
   trick").

   The second disadvantage of the PCA consists in the fact that the
   directions maximizing variance do not always maximize information. The
   page of the LDA subroutines gives an example of such a task, wherein the
   maximum-variance variable affords almost no information, whilst the
   minimum-variance variable permits classes to be wholly separated. In this
   case, the PCA will give preference to the first (less informative)
   variable. This drawback is closely connected to the fact that the PCA
   does not perform linear separation of classes, linear regression or other
   similar operations, but it merely permits the input vector to be best
   restored on the basis of the partial information about it. All additional
   information pertaining to the vector (such as the identification of an
   image with one of the classes) is ignored."

   Based on this description, and if this description is correct, I believe
   that the authors should provide more concrete evidence showing that such
   assumptions are met in their analysis, so that the authors can make some
   inferences from the PCA results. However, it is unclear to me based on
   the current manuscript.

   In addition, I found that it is unclear what are the take-home messages
   from the current analysis. It sounds that more research should be done.
   If that is the case, this study sounds like a quite exploratory at this
   stage.

   * page 2
   “speed[9]”

   Put a space right before a citation throughout this paper.

   * page 3
   "Lineups require similar manipulations in two-dimensional space, and also
   require the ability to perform complex spatial manipulations mentally;
   for instance, comparing the interquartile range of two box- plots as well
   as their relative alignment to a similar set of two boxplots in another
   panel."

   I understand that the experiment follows the line-up protocols, but it is
   unclear how it was conducted in detail. What are the visualizations used
   for the three blocks of 20 lineups. What are tasks? (Task examples are
   shown in Figure 1(a), but not sure if the same task was used for all
   lineups). I later found that more examples of line-up tasks are described
   in Section 3.6, but the detailed descriptions of tasks are still
   necessary.

   * page 6
   "3.5 Lineup Types"

   This should be go to the Methods section, not the result section.

   ---

   Regarding the concern regarding PCA, a reviewer who had the sufficient
   statistics background mentioned that my concern is irrelevant to the
   application in the paper. I hope that the authors clarify potential
   questions from readers by clarifying why PCA would be an appropriate
   approach for the people who do not have sufficient statistical
   background.

# Review 1

_Title_: Spatial Reasoning and Data Displays

_Reviewer_:           external

_Paper type_   Evaluation

_Expertise_ 2  (Knowledgeable)

_Overall Rating_ 3.5 - Between Possible Accept and Accept

_Supplemental Materials_  Not applicable (no supplemental materials were submitted with the paper)

## Justification

The paper addresses an important issue in spatial reasoning on graphics.
   While the paper is well presented and the study is extensively evaluated,
   I find that the paper could benefit from stronger motivation of certain
   design choices, better explanation of experimental procedures, and more
   conclusive evaluation of the results.

## The Review

   The paper addresses an important issue in spatial reasoning on graphics
   that I believe is highly valuable to the InfoVis community. The paper is
   well presented and the study is extensively evaluated.

   My main concerns about the paper are threefold.

   Firstly, the paper could benefit from stronger motivation of certain
   design choices. For instance, it is not fully clear how the measures of
   visuospatial ability were chosen from a much larger set of tests in the
   given reference. How are these related to spatial and reasoning ability
   and what were the hypotheses about their relation to the graphics
   presented in the lineup tasks?

   Secondly, the descriptions of the experimental procedures is lacking
   detail. In 2.2 the different tests are explained but it is not clear in
   which order these were presented. Was it a randomised or counterbalanced
   design? Also, what was contained in the 20 line ups presented between
   each session in relation to the different plot types presented in Fig 9?
   Why were the lineups interleaved with the other four tests? What was the
   experimental setup: screen used, viewing distance, lab environment? What
   were the instructions for each of the experimental sessions? What was the
   total experiment time and were breaks included? Beware that such studies
   can be influenced by many factors, such as order effects, fatigue, and
   learning effects, for which reason it is really important to provide
   sufficient details on the experimental design.

   Lastly, while the evaluation appears to be very extensive, I find the
   results not as conclusive as they are interpreted by the authors. Some
   rather low correlations around 0.3-0.4 are interpreted by the authors as
   being high, which I would not agree with. For instance, in Fig 10
   differences in correlation between 0.2 and 0.3 are interpreted to be
   significant while the respective scatter plots do not suggest for one or
   the other to be significantly better. Such small differences in low
   correlation regimes can often be misleading and should not be
   overinterpreted. I appreciate the authors' effort to also draw strong
   conclusions on the results of the PCA, but also these appear to be not
   overly convincing. For instance, at the end of 3.5 the author state that
   lineup 3 is different from lineup 1 and 2. I agree with that from visual
   observation in Fig 9, but I would be hesitant to infer that from the PCA
   results.

   Some more detailed and minor comments:
   - In 2.2, Wolfe's work on visual search should be mentioned in relation
   to the VST.
   - Equation (2) states the same as equation (1).
   - In section 3, you mention one exception in relation to the
   participants' age, please be more specific.
   - In section 3, I suppose you mean that the demographics are
   representative for the student population and not for the whole
   university incl staff.
   - With regard to your assessment of the correlation in Fig 6. I would
   discriminate more between the individual results rather than stating that
   they are all highly correlated.
   - It would be good to move Table 1 to the previous page.
   - Concerning Fig 5: I am not surprised that there is no difference for
   age, as the difference is minimal and the groups would probably be put
   into the same group would there be a wider spread of age in the
   participant population.
   - In the last paragraph of section 3.3: There is a typo "... is
   ultimately are a classification... ".
   - Beginning of 3.4: you mention that all demographic variables are highly
   correlated. Really all? Please be more specific.
   - Task for lineup 3 is given. What were the tasks given for the
   evaluation of lineups 1 and 2?
   - The VST task is excluded from Fig 10 due to its low correlation. Please
   include it for completeness.
   - PCA is explained in detail in 3.5 although it has been used extensively
   earlier. Please put the explanation before the first use.
   - Section 4: I don't agree with the initial statement that lineups are
   strongly related to the performance on tests of visual ability. There is
   certainly evidence provided for that in the paper, but I would not refer
   to it as strong evidence.
   - I propose to put Appendix C into the main body of the paper.
   - In Appendix C: I don't agree with the statement that the card rotation
   task is much more associated with the QQ plots. The correlations are
   actually higher for the first four plot types on the left in the figure.

# Review 3

_Title_: Spatial Reasoning and Data Displays

_Reviewer_:           external

_Paper type_  Evaluation

_Expertise_ 2  (Knowledgeable)

_Overall Rating_ 2.5 - Between Reject and Possible Accept

_Supplemental Materials_ Acceptable

## Justification

   This paper pairs visualization judgments with cognitive tests to try to
   disentangle cognitive strategies for asserting the "most different" plot
   in a set. The paper asks an interesting question and is very well
   written. However, it is unclear from the manuscript what the experimental
   task is and how it's controlled for, and the analysis is of questionable
   validity as it aggregates results across a wide breadth of individual
   designs without consideration for how different visualization designs
   might lead to different cognitive strategies. As a result, the paper is a
   near-miss: it asks an interesting question, but makes a few very critical
   oversights that make its results both questionable and of little utility
   for the visualization community.

## The Review

   The goal of this paper is to measure the role of different forms of
   spatial reasoning in interpreting visualizations, with the ultimate goal
   of understanding the visual strategies underlying judgments made using a
   visualization. Turning to cognition to try to illuminate different
   strategies for interpreting visualizations is an extremely interesting
   idea. However, there are several elements of the experimental design and
   analysis that are concerning. Of particular concern are the ambiguity of
   the lineup task and the lack of consideration of different design
   elements within the analysis. While I believe that the paper in its
   current form is unsuitable for InfoVis, the manuscript is headed in the
   right direction and could be a significant contribution to the community
   in the future.

   It is unclear in the experimental task what “most different” means.
   From the text, it seems as if its intended to measure distributional
   variation. However, “most different” can (and likely will) be
   interpreted by participants in different ways (e.g. largest difference in
   variance, largest difference in mean or median). Was this communicated to
   the participants? Were they given a tutorial? How was the difficulty of
   this judgment controlled? How were other potentially related statistics
   controlled for (e.g. how can we verify that participants were actually
   performing the requested task and not a related proxy task)? What
   information was available to participants for each visualization design
   to make this judgment? In a box plot, for example, we’re only getting
   distributional characteristics, it’s possible that the box plot is
   insufficient to characterize the distribution effectively (e.g.
   Anscombe’s quartet).

   A larger issue related to this point is that the manuscript lacks a
   discussion of visualization design in general. For example, the
   manuscript clearly indicates that  there is a difference in the reported
   correlations across different sets of visualizations (e.g. lineup one,
   two, and three). This directly suggests that the different visualization
   design components , which is well-explored in the graphical perception
   literature (e.g. Cleveland & McGill, 1984, Heer, Kong, & Agrawala, 2009,
   Fuchs et al, 2013, Albers, Correll, & Gleicher, 2014 among many others).
   This would also imply that the findings as reported are biased by the
   sampling of visualization types used in the experiment. Further, the
   comparison within lineups seems somewhat unprincipled. For example,
   it’s unclear how a dot-plot is more similar to set one than set two.

   This will likely be a significant confound in the comparisons with the
   more traditional  cognitive tests. Assuming the tests use exactly the
   same form of stimuli founding figures 2 and 3, these stimuli are uniform
   in design elements across all of the visualization designs. As a result,
   it’s difficult to tell if the results are a by-product of similarity in
   visual features or in visual strategy. This analysis feels like it might
   be comparing apples to oranges--I am not convinced that it makes sense to
   compare the lineup results to performance on these tests in aggregate due
   to complications in visual design. Given this confound, it is unclear how
   a visualization designer could use these results in practice.

   For example, the lineup test is an example of a more free-form visual
   search task: participants are actively searching for the “most
   different” element. As a result, deviations from these two tasks might
   be a function of difference in strategy (as suggested in the paper) or
   due to difference in stimuli (e.g. detecting configurations vs. box
   plots) or both. The figure classification test may provide the most
   similar results for many visualization designs simply because the visual
   hull of the data (the “cloud” in a scatterplot, the box in a boxplot)
   is the most salient global feature for making similarity comparisons,
   which aligns well with the "form" of a figure. This would lead to
   correlation between the tests, but the actual strategy may not be
   rule-based.

   To make these comparisons more compelling, these same experimental
   procedures can be conducted with the stimuli in each of these tests
   adapted to more closely mirror the tested visualization type. I would
   encourage the authors to consider restructuring these tests to use
   stimuli more relevant to each respective visualization lineup condition
   or, at a minimum, to at least break down the results with respect to each
   individual visualization design and compare strategies across visual
   feature. Aggregating across all visualizations is an oversimplification
   of the task—different visualization may require different strategies,
   and, in fact, graphical perception suggests that they do.

   I would also recommend that a revised manuscript include an explicit
   related works section. Traditionally in this community, the related work
   is separate from the introduction. While I think this paper reads well
   without that structure, the authors may wish to consider revising their
   structure to accommodate a more thorough discussion of related work. Part
   of this recommendation is that there is a good breadth of relevant work
   that is not considered in this manuscript. For example, “The lineup
   protocol [12, 25, 3] is a testing framework that allows researchers to
   quantify the statistical significance of a graphical finding with the
   same mathematical rigor as conventional hypothesis tests” is a very
   contentious claim. In particular, recent work in visualization, such as
   (Pak, Hutchinson, & Turk-Brown, 2014, Talbot, Setlur, & Anand, 2014 (more
   distant comparison is harder), Correll & Gleicher, 2014, and a myriad of
   papers on limitations in interpreting boxplots), suggest that people are
   not good at performing visual significance testing. Visual search is
   among the most commonly used tasks in graphical perception experiments.
   For example, Fuchs et al, 2014 includes experiments that explicitly
   involve visual search tasks in small multiple (lineup) conditions.

   Minor Issues:
   Additional experimental details about the type of visualizations tested
   (e.g. Figure 9) should be presented in the Methods section. A pointer to
   another set of papers with large numbers of potential designs is
   insufficient to understand the experiment. Additionally, explanations of
   the baseline cognitive tests should be presented up front to ground the
   analysis: while many in the visualization community may have seen the
   compared tests (visual search task (VST), paper folding test, card
   rotation test, and figure classification test), especially the visual
   search task, many other have not. Give the necessary grounding and
   explain why these particular tasks were chosen as a baseline.

## Spelling & Grammar:

   Overall this paper is very well written. However, throughout the paper,
   the term “graphics” is used instead of “visualization”.
   “Graphics” has a much broader possible meaning and
   “visualization” is conventional for the InfoVis audience.

   speed[9]  -> speed [9]
   reasoning[6] -> reasoning [6]

# Review 4

_Title_: Spatial Reasoning and Data Displays

_Reviewer_:           secondary

_Paper type_   Theory / Model

_Expertise_   3  (Expert)

_Overall Rating_ 5 - Strong Accept    
The paper is an excellent contribution and should certainly be accepted, possibly with some minor revisions.    
It more than meets the quality standard.

_Supplemental Materials_ Acceptable

## Justification

   Nothing that I haven't listed in the review. I think this is an
   extraordinary paper because the authors have taken a fundamentally
   important contribution and extended it to an area where nobody has
   thought about it so far.

## The Review

   The work represented here may be the most important recent advance in
   methodology for evaluating the effectiveness of visual displays. It
   appears to be, as well, the most powerful method for evaluating
   individual differences in graph reading ability. What is remarkable is
   that such a simple paradigm can be coupled with a formal statistical
   distribution (the binomial) for quantifying these results. Instead of
   separating graph reading into subtasks that may be highly correlated (and
   therefore not highly informative taken as a whole set) this team has
   devised a method for dealing with highly complex displays (probability
   plots, billets, etc.) as well as simple ones.

### 2015-06-08 Paper/InfoVis2015/ReviewerResponses.Rmd

   Moreover, this team has implemented a methodology that fits perfectly
   into a tradition of careful psychological measurement. Instead of
   requesting subjects to rate things numerically (as Cleveland did, for
   example), these researchers simply ask for the point of a finger, so to
   speak. The response is similar in nature to that of classical perceptual
   direct scaling studies going back almost a century.

   Sometimes authors or teams milk a simple idea by generating minor
   variations and submitting them to a bunch of journals. What's remarkable
   here is that the collection of their publications given in the references
   ranges from a formal statistical article with proofs etc. to several
   visualization articles having very different focuses. The same is true
   for this article. I have seen presentations of this research and I never
   thought of the possibility that it could be used in the context of the
   present article.

   This paper is very well organized and written. I have few considerations
   for revision and plan to nominate it for best paper.

### Details:

   I'd suggest a mention of the groundbreaking mental rotation research of
   Shepard & Metzler. This was 3D rotation, but it did stimulate thinking by
   later psychologists on the rotation problem. What's more, their results
   were confirmed by subsequent neurological research in Science magazine.

### 2015-06-08 Paper/InfoVis2015/TestingVisualAptitude.Rnw
Reviewer responses and slightly modified regex in paper
Modifications - "most different" = feature, not bug; PCA discussion + additional reference

sanitize <- function(x){
  str_replace(gsub("_", ".", x, fixed=T), "Pr\\((.{1,}))", "Pr$(\\1)$")
}

@@

Results from the visuospatial tests used in this experiment are highly correlated, as shown in Figure~\ref{fig:scatterplotmatrix}; this is to be expected given that all of these tests are in some way measuring individuals' visual ability. 
What is of more interest to us is how other factors, such as e.g.~general intelligence, mental processing speed, cognitive resources, motivation, and attention affect performance. 
In order to assess factors contributing to lineup performance, we first examine the separate dimensions measured by the battery of cognitive tests (other than lineups) using principal components analysis on the scaled test scores, then we examine all five tests using the same procedure.

A principal component analysis (PCA) of the four established visuo-spatial tests reveals that they all share a very strong first component, which explains about \Sexpr{round(summary(pca2)$importance[3,1],2)*100}\% of the total variability. Principal components (PC-) are ordered by importance (how much variability in the data they contain), and each principal component is uncorrelated with every other principal component.

>>>>

sanitize <- function(x){
  str_replace(gsub("_", ".", x, fixed=T), "Pr\\((.{1,})\\)", "Pr$(\\1)$")
}

@@

Results from the visuospatial tests used in this experiment are highly correlated, as shown in Figure~\ref{fig:scatterplotmatrix}; this is to be expected given that all of these tests are in some way measuring individuals' visual ability. 
\newtext{While test scores are correlated, the relationships between test scores appear to be linear, with sufficient variability around the trend line to merit further investigation. The correlations are to be expected, as all of the tests in this experiment are conducted in a visual domain, and require general reasoning, intelligence, and other factors. }
What is of more interest to us is 
\newtext{whether lineups recruit specific cognitive abilities, such as visual search, reasoning and categorization ability, mental rotation.} 
In order to assess factors contributing to lineup performance, we first examine the separate dimensions measured by the battery of cognitive tests (other than lineups) using principal components analysis on the scaled test scores, then we examine all five tests (four cognitive tests, and additionally lineups) using the same procedure.

\newtext{
Principal component analysis (PCA) is a statistical technique to transform the coordinate system of correlated data, producing orthogonal components in a new coordinate space (see \cite{pcaReference} for more information). The most stringent assumption required for PCA is that of linearity, which is reasonable given Figure \ref{fig:scatterplotmatrix}. Additionally, PCA assumes that the interesting portion of the data is the variance between any two variables (a perfectly linear relationship between two variables would be uninteresting, as the variables contain the same information). As a result, PCA can be used to examine which variables contain similar information (i.e. which variables are represented in each principal component). It is this feature of principal components which we utilize in this paper. 
}

A principal component analysis of the four established visuo-spatial tests reveals that they all share a very strong first component, which explains about \Sexpr{round(summary(pca2)$importance[3,1],2)*100}\% of the total variability. Principal components (PC-) are ordered by importance (how much variability in the data they contain), and each principal component is uncorrelated with every other principal component.
